Discussion Draft

### 2020-03-25 doc/Discussion.rmd
Interference parts

@@

The strongest correlate of faster gwRR in males was intercrease interference strength. This is unexpected beacuse most literature and logical models assoicate more dense spacing of crossovers along chromosomes. Luckily having unexpected results is good for distingushing the most parimonius model. These results were unexcepted and differ ffrom the major patterns listed in the literature (otto payseur, argawal Korol. (it also goes against the logical assumptions models of fitting more crossovers along the SC chromosome axis space)

**The rapid evolution of the genome wide rate in our high rec strains is PARTLY due to more enrichment of 2CO which have distinct rec landscape from the 2COs in low rec strains.**

We considered carefully if this pattern was due to error -- and we are confident that this pattern is genuine -- goes aginst expected noise ALSO this is observed from other work in this lab -- and some empiracal studies (cattle) (plant bomblies).

<while these results don't clearly fit the chromatin organization and interference strength prediction... determining the clean pattern is beyond the scope of this paper>

HOW do the models above treat this unexpected correlation?

**typical modifer models -- are not good for predicting interferece (in mode that is relevant to these types of results) since the models don't parameterize for indiviual meiosis events.** We conclude that under the haploid selection and meiotic driver model any prediction for evolution of interference strength follows from the 'logical' expectation of a negative correlation.

originally proposed/outlined by Trivers [] and refined /formatlized in [@lenormand2003]. This model discribes a system where gametes are competing or selection is acting in the haploid stage to cause difference in genome wide recombiaion rates between males and females.

 This is a modefier model of three loci, where the recombination rate between two loci is determined by a third locus (or modifer).

@@

**spindle based selection model**
There is a model that predicts this positive ocrrelation. The goldstein results (reviewed in ottoPayseur). when the number of crossovers (the obligate crossover assumption is more strict / the
breifly, the model simulates / sets up a scnario where a modifer acts through the probabilty of flanking regions having 0 or 2 crossovers. ith crossover assuance, recombination rate increases the further apart the loci are spaced (?)

Additionally there are a handful of empircal results which also show the unexpected pssotive correlation between gwRR and interference strength.

1. Cattle (Ma?) [@ma2015_cattle] ? (or newer one), when comparisons of genetic map length and interference strength are made across strengths within sex, the strain with the higher genome wide recombinaiton rate also has stronger interference (measured as nu gamma parameter)

2. The IFDs from (Beth's cross) a related house mouse cross show that the PWD (same strain, different mice) had longer IFD compared to Cast (HVR unpublished data). (also gamma?)

3. In Peromyscus, the wild mice had higher gwRR and stronger (gamma nu estimates) compared to lab raised mice of the same speicies [@peterson2019]

<note about the differences in the patterns>  assumptions with gamma dstribution and linkage maps, (closer to infinte genetic space), less weight on crossover assurance - more weigth on 
(pooled SC/chromosome length gives strong correlation )

<which is much different than when a single chromosome is examined (by itself), finite 2D length, crossover assurance, weaker correlation with ((the closer to chromosome level for SC legnth metric / summary, the weaker the correlation with crossover number))

@@

4. HYPER-Crossover experiments. (review the effects when crossover number is artificially altered)

This new model has **strengths**

  - It matches fine scale differences in how DSB hotspots had different sex specific strengths with varied due to sex, epigenetc landscape and the gene architecture (Brick et al)

@@

- extends modifer model to explain a chromosome based / level pattern

_weaknesses_, and ways the model diverges from our results

 - the simulation present in the paper are too weak to support the rapid spread of such a modifer system (selection coefficients are too weak to support the evolution -- spread of such a modifer.)
- Other ways this model departs from our results, 
-Always predicting male will be lower recombination rates (or risk breaking up the epistatsis blocks) -- which doesn't fit with our results of faster male specific evolution in gwRR.
However / while not explicitaly arameterized or predicted in this model,
 -- this model predicts larger chuncks of chromosomes will segregate together
 (as outlined in Veller et al)

 our results of stronger male interference and stronger interference in high rec strains have the same consequences as predited from this modle

@@

# The CONTEXT of MEIOSIS

the study confirmed preivously identified sexual dimorhisc patterns in regard to the recombiantion landscape (in mammals) -- in the goal of unifying all of these results, we sough to investigate physiologycal aspects of gametogenesis which have conserved sexual dimorphic features.

A fundamental difference in meiosis (MI) for males and females is the presence of the centrosome which serves as a nucleation site for MT at each pole (and changes the shape of the spindle). (Our hypothesis is that the tension in a centrosome spindle is stronger (or more uniform) since all MT-KT are anchored at single points across cells (the two centrosomes). This will also make prediction for the spindle assembly checkpoint (SAC), which is stronger/stricter in smaller gametes with centrosomes.  (the SAC is more sensitive to achiasmate bivalents on the spindle / better at detecting a lack of tension).
Whereas in the acentrosome spindle is more diffuse (across a larger area) and the tension across bivalents are anchored by multiple MTOCs, resulting in weaker tension force or less uniform strengths across all the bivalents within the cells. The SAC in eggs seems to be weaker/leakier, that is more achiasmate bivalents are required to trigger the SAC (stop the division).

_main sex differences in spindle_ are the anchoring of MT at opposite poles

@@

# CO placement translates to bivalent shape

@@

# Future Steps

be aware of sex-specific and sex average biases

- linkage / R between two foci is sex averaged

- Celegans -- cytlogy focuses on oogenesis  and genetic maps are often from hermaphrodites (selfing -- sperm and egg landscapes are merged in hemaphordites)

-- the value of Chromosome level measures -- incorporate chromosome behavior into models

-  more attention to cell physiology and biophysics (conserved nature of meiosis pathway (and mitosis ) -- use some models such as Sc for manipulation) (Murkarjee gardner -- quantify the spindle tension

-- lack of SC length prophase cell charaecterization
-- comparisons within and between sexes (that cattle paper)

# Reference

@@

However, this model doesn't explain the borad scale pattern for the male pattern.

**weaknesses**  (modifers - driver systems everywhere!), impossible to verify, doesn't account for male landscape pattern (supression from centromere to end)

this model would predict / fit the results of>

<**model basics**, 'break up driv system consiracies'>

@@

>>>>

#**The rapid evolution of the genome wide rate in our high rec strains is PARTLY due to more enrichment of 2CO which have distinct rec landscape from the 2COs in low rec strains.**

@@

Our results indecated that interference strength has evolved with the genome wide recombiantion rates in the three strains with faster male evolution.

 This is unexpected because most literature and logical models are built / assume that weaker dense spacing of crossovers along chromosomes (ie. weaker interference) is how .

The history, models and empiracal examples are extensively reviewed in [@ottoPaysuer2019]. It is notable that borad discriptions / broad scale patterns for the correlation of gwRR and interference strength (gamma parameter) do fit the expected negative pattern [@ottoPaysuer2019].

(genome wide rate vs single bivalent measures)

**Caveats**
We considered carefully if this pattern was due to error and we are confident that this pattern is genuine.

There's no model that fits evolution of heterochiasmy and interference independantly. Under our undertsanding of the models, the relatioship of genome wide recombiantion rate and interference strength is assumed to be negative. We now review how some of the assumptions of typical modifer models might not fit realities of the meiotic pathway.

First, the **typical modifer models do not model single meiotic events**. (think about how pooling man meiotic events can skew patterns)
(crossover interference can be quantified from linkage maps (which are the average of many hundreds to thousands of meiotic events). It can also be discribed / measrued /quantified from single bivalets / chromosomes (either cytology of dense genotyping of single cells (cite). 
While it's widely agreed that they both reflect the same biological process, these two approaches are fundamentally different. (they are subject to distinct biases and noise).

Second, most modifer models use abstracted genomic space -- like an infite 2D line between markers. (space between markers is abstracted). This is useful for fitting the gamma model, but within the cell, the physical chromosomes's axis length is fitnite and crossovers cannot form beyond those bounds.

We conclude that under the haploid selection and meiotic driver model any prediction for evolution of interference strength follows from the 'logical' expectation of a negative correlation.

originally proposed/outlined by Trivers [] and refined /formalized in [@lenormand2003]. This model describes a system where gametes are competing or selection is acting in the haploid stage to cause difference in genome wide recombination rates between males and females.

 This is a modifier model of three loci, where the recombination rate between two loci is determined by a third locus (or modifier).

@@

**spindle based selection model**
There is a model that predicts this positive correlation between genome wide recombination rate and interference. The goldstein results (reviewed in).

The results from golstein reviewed in [@ottoPaysuer2019], predict a modeifier model (which placed more weight on crossover assurance (holds the number of crossovers constant) and simulates a scnario where a modifer acts on modulating the probabilty of flanking regions having 0 or 2 crossovers. Which has the result of increasing the distance between two foci.

Additionally there are a handful of empircal results which also show the unexpected possitive correlation between gwRR and interference strength. When comparisons of genetic map length and interference strength are made across strengths within sex, the strain with the higher genome wide recombinaiton rate also has stronger interference (measured as nu gamma parameter) [@ma2015_cattle]. In *Peromyscus leucopus*, the wild mice had higher gwRR and stronger (gamma nu estimates) compared to lab raised mice of the same speicies [@peterson2019].

Additionaly -- preliminary data from a previous cross using MLH1 cytology in spermatocytes in house mouse replicated these results with IFD measures from  M. m. musculus^PWD^  being longer compared to M. m. castenus^Cast^ (HVR unpublished data) (cast has low gwRR).

note about the differences in the patterns>  assumptions with gamma dstribution and linkage maps, (closer to infinte genetic space), less weight on crossover assurance - more weigth on 
(pooled SC/chromosome length gives strong correlation )

which is much different than when a single chromosome is examined (by itself), finite 2D length, crossover assurance, weaker correlation with ((the closer to chromosome level for SC legnth metric / summary, the weaker the correlation with crossover number))

@@

This new model has **strengths**:
  - It matches fine scale differences in how DSB hotspots had different sex specific strengths with varied due to sex, epigenetc landscape and the gene architecture (Brick et al)

@@

- extends modifer model to explain a chromosome based / level pattern
 _weaknesses_, and ways the model diverges from our results

 - the simulation present in the paper are too weak to support the rapid spread of such a modifer system (selection coefficients are too weak to support the evolution -- spread of such a modifer.)
- Other ways this model departs from our results, 
-Always predicting male will be lower recombination rates (or risk breaking up the epistatsis blocks) -- which doesn't fit with our results of faster male specific evolution in gwRR.
However / while not explicitaly arameterized or predicted in this model,
 -- this model predicts larger chuncks of chromosomes will segregate together
 (as outlined in Veller et al)
our results of stronger male interference and stronger interference in high rec strains have the same consequences as predited from this modle

@@

## The CONTEXT of MEIOSIS

the study confirmed preivously identified sexual dimorhisc patterns in regard to the recombiantion landscape (in mammals) -- in the goal of unifying all of these results, we sough to investigate physiologycal aspects of gametogenesis which have conserved sexual dimorphic features. A fundamental difference in meiosis (MI) for males and females is the presence of the centrosome which serves as a nucleation site for MT at each pole (and changes the shape of the spindle). (Our hypothesis is that the tension in a centrosome spindle is stronger (or more uniform) since all MT-KT are anchored at single points across cells (the two centrosomes). This will also make prediction for the spindle assembly checkpoint (SAC), which is stronger/stricter in smaller gametes with centrosomes.  (the SAC is more sensitive to achiasmate bivalents on the spindle / better at detecting a lack of tension).
Whereas in the acentrosome spindle is more diffuse (across a larger area) and the tension across bivalents are anchored by multiple MTOCs, resulting in weaker tension force or less uniform strengths across all the bivalents within the cells. The SAC in eggs seems to be weaker/leakier, that is more achiasmate bivalents are required to trigger the SAC (stop the division).

**main sex differences in spindle** are the anchoring of MT at opposite poles

@@

#CO placement translates to bivalent shape

@@

#Future Steps

Be aware of sex-specific and sex average biases. Some cytlogy models (Celegans) mostly focus on oogenesis  and genetic maps are often from hermaphrodites. (lack of data from one sex). Recognize the value of Chromosome level measures and incorporating chromosome behavior into models.

Borader communication across sub-fields of cell physiology and biophysics for understanding recombination rate evolution. Utilize mitosis systems given the conservation of the molecular pathway. THink of meiosis as a program.

Dark matter of the genome. We can't really explude that some of the recombination rate variation is due to strucutreal variation (more of less genome). (since everything is mapped to B6). (also maybe there are crossovers in the centromeres and telomeres -- which might escape detected in linkage maps)

what are the rates of sister-sister crossovers?

How does elevated rates of precocious sister centromere seperation effect oogenesis?

Are selfish elements affecting the evolution of meiosis -- and thus recombination rates (from a functional standpoint).

Keep generating those testable hypotheses (cite).

#blah

@@

However, this model doesn't explain the borad scale pattern for the male pattern.>

@@

#References
